Estimating duration in the presence of misclassification

### 2024-08-02 cis-imperfect-testing.tex
Updates from Overleaf

In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of detectability was developed.
However, the initial application of this framework produced an implausibly short estimate.
In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling}).
After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.

@@

The CIS data provide evidence for the test sensitivity declining over the course of an episode.
\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
\begin{enumerate}
    \item All positive results are true positives.
    \item All intermittent negatives are false negatives.
    \item The negative at the end of an episode is a possible false negative.
    \item All other negatives are true negatives.
\end{enumerate}
The classification assumes that the probability of two false negative tests at the end of an episode is negligible.
As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
This can be bounded by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
I calculate the test sensitivity for each day, where the day of the first positive result in an episode is day zero.
Day zero is an unknown number of days after the episode's latent true beginning time.
% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie the first positive test in the episode).
I generate broad bounds, suggestive of a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
However, the general trend is clear.
\begin{figure}
  \includegraphics{cis-imperfect-testing/test-sens-bound}
  \caption[Bounding test sensitivity using CIS data]{
    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
  }
  \label{imperf-test:fig:bounding-cis-sensitivity}
\end{figure}

@@

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}; I continue to assume the probability of a false positive is negligible.
Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.
Additionally, I limit the number of permutations of false negative and true positive tests that I consider have non-negligible probability.
Specifically, I assume that, if $i(j)$ is tested during infection episode $j$, either: the individual has a single false negative test and the episode ends before their next test, or the first test is a false positive test.
This assumption is reasonable because false negatives normally occur when the viral load is low, which is normally late in the infection episode.
For the same reason, I assume there is at most one false negative test following the final true positive test in the episode.

@@

Introducing false negatives means that $O(j)$, containing the relevant information for episode $j$, is now random, even if $B_j$ and $E_j$ are known.
Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
For tractability, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.

To do so, define a random vector $O'(j)$ with state space $\{\emptyset\} \cup \set{E}'$ which will replace $O(j)$ from \cref{E-perf-test}.
As before, $O'(j) = \emptyset$ if episode $j$ is undetected.
Otherwise, $O'(j) \in \set{E}'$ where $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ augments the space $\set{E}$ with test results during the episode.
The elements of $\set{E}'$ are all $\vec{\nu}_k' = [\vec{\nu}_k, \vec{y}_k]^T$ where $\vec{\nu}_k \in \set{E}$ and $\vec{y}_k$ is a vector of test results for the relevant testing times, $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$.
To formalise the definition of $\vec{y}_k$, let $m_k$ denote the size of $\sched'_k$ and denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
Then, $\vec{y}_k \in \{0, 1\}^{m_k}$ and satisfies the following two conditions.
\begin{enumerate}
  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ is negative, \ie $y_{k,m_k} = 0$.
\end{enumerate}
These conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.

Let $p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
As before, $p'_k = 1/\Ncis p'_{ik}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'_{iu}$.

@@

By assumption, the negative test bounding the start of the episode, on day $l_k^{(b)}-1$, is a true negative.
True negatives occur with probability 1, and hence this test does not contribute to the likelihood.

@@

For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
By assumption, there is a negligible probability of two false negatives.
Therefore, this can only occur if the episode ends before another test.
Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j)$.

>>>>

In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of detectability was developed.
However, the initial application of this framework produced an implausibly short estimate.
In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.

@@

The CIS data provides evidence for the test sensitivity declining over the course of an episode.
\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
\begin{enumerate}
    \item All positive results are true positives.
    \item All intermittent negatives are false negatives.
    \item The negative at the end of an episode is a possible false negative.
    \item All other negatives are true negatives.
\end{enumerate}
The classification assumes that the probability of two false negative tests at the end of an episode is negligible.
As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
This can be bounded by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
I calculate the test sensitivity for each day, where the day of the first positive result in an episode is day zero.
Day zero is an unknown number of days after the episode's latent true beginning time.
% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie the first positive test in the episode).
I generate broad bounds, suggestive of a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
However, the general trend is clear.
\begin{figure}
  \includegraphics{cis-imperfect-testing/test-sens-bound}
  \caption[Bounding test sensitivity using CIS data]{
    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
  }
  \label{imperf-test:fig:bounding-cis-sensitivity}
\end{figure}

@@

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.

@@

Introducing false negatives means that $O(j)$, the observation vector of episode $j$, is now random, even if $B_j$ and $E_j$ are known.
Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
For tractability, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.

To formalise this, I define analogous variables to those in \cref{E-perf-test:sec:posterior}.
To start with, define the relevant testing times for any episode $j$ with $O(j) = \vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
Denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
Let $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ be a new set of possible observations of detected episodes, including the pattern of positive and negative tests.
Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_k \in \{ 0, 1 \}^{m_k}$ with $y_{k,l}$ being the test result at time $t_{k,l}$.

$\vec{\nu}'_k \in \set{E}'$ occurs if and only if all the following conditions are met.
\begin{enumerate}
  \item $\vec{\nu}_k \in \set{E}$.
  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ is negative, \ie $y_{k,m_k} = 0$.
\end{enumerate}
The final two conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.

Let $O'(j)$ be the observations generated by episode $j$ including the new vector of results.
As before, $O'(j) \in \set{E}'$ indicates that episode $j$ was detected; otherwise, $O'(j) = \emptyset$.

Similarly, let $p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ have the same definition as p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
As before, $p'_k = 1/\Ncis p'_{ik}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'_{iu}$.

@@

For tractability, assume that the negative test bounding the start of the episode, on day $l_k^{(b)}-1$, is a true negative.
This assumption is reasonable because, since a positive test follows at $r_k^{(b)}$, the negative at $l_k^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
True negatives occur with probability 1, and hence this test does not contribute to the likelihood.

@@

For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
I assume there is a negligible probability of two false negatives.
Therefore, this can only occur if the episode ends before another test.
Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j)$.
